---
title: "Analysis Notebook"
toc: true
toc-depth: 3
number-sections: true
fig-width: 8
fig-height: 6

execute:
  freeze: false
  cache: false

---

# Setup
```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(emmeans)
  library(DescTools)
  library(coin)
  library(XNomial)
  library(brms)
  library(lme4)
  library(multcompView)
  library(patchwork)
  library(tidyverse)
  library(readxl)
})

# Set paths
exc_file_dir    <- "data/raw/Mesocosm_experiment.xlsx"
raw_data_dir    <- "data/raw"
der_data_dir    <- "data/derived"
out_dir         <- "outputs"

read_sheet <- function(sheet) {readxl::read_excel(exc_file_dir, sheet = sheet)}
write_out <- function(df, name) {write.csv(df, file.path(der_data_dir, name), row.names = FALSE)}

# Read data
df_koura               <- read_sheet("Koura")
df_setup               <- read_sheet("Tank_setup")
df_exp1_individual_raw <- read_sheet("exp1")
df_exp2_group_raw      <- read_sheet("exp2")
df_exp3_woodstone_raw  <- read_sheet("exp3")
df_exp4_material_raw   <- read_sheet("exp4")
df_wq                  <- read_sheet("WQ")
df_loggers             <- read_sheet("Loggers")

# Save as csv to derived
write_out(df_koura,              "df_koura.csv")
write_out(df_setup,              "df_setup.csv")
write_out(df_exp1_individual_raw,"df_exp1_individual_raw.csv")
write_out(df_exp2_group_raw,     "df_exp2_group_raw.csv")
write_out(df_exp3_woodstone_raw, "df_exp3_woodstone_raw.csv")
write_out(df_exp4_material_raw,  "df_exp4_material_raw.csv")
write_out(df_wq,                 "df_wq.csv")
write_out(df_loggers,            "df_loggers.csv")

# set preferred orders
STRUCTURE_LEVELS  <- c("SingleStone","FlatStone","StonePile","WoodLog","TankWall")
STRUCTURE_LEVELS3 <- c("FlatStone","WoodSplit","TankWall")
STRUCTURE_LEVELS4 <- c("ConcreteBrick","WoodBlock","TankWall")
LOC_LEVELS   <- c("A", "B", "C", "D", "E")
LOC_LEVELS3  <- c("A", "B", "E")

relevel_structure  <- function(df) df %>% mutate(structure_type = factor(as.character(structure_type),levels = STRUCTURE_LEVELS))
relevel_structure3 <- function(df) df %>% mutate(structure_type = factor(as.character(structure_type),levels = STRUCTURE_LEVELS3))
relevel_structure4 <- function(df) df %>% mutate(structure_type = factor(as.character(structure_type),levels = STRUCTURE_LEVELS4))

# Set plot theme
windowsFonts(Arial = windowsFont("Arial"))
base_theme_bw <- theme_classic() +
  theme(text     = element_text(family = "Arial", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))
theme_set(base_theme_bw)

get_sex <- function(tbl, sex, col) {
  tbl[[col]][tbl$sex == sex]
}

get_size <- function(tbl, class, col) {
  tbl[[col]][tbl$size_class == class]
}

# Build pairwise_gtest
pairwise_gtest <- function(counts) {
  nms   <- names(counts)
  pairs <- combn(nms, 2, simplify = FALSE)
  
  results <- lapply(pairs, function(p) {
    obs_pair <- counts[p]
    exp_pair <- rep(sum(obs_pair) / 2, 2)
    
    G     <- 2 * sum(obs_pair * log(obs_pair / exp_pair), na.rm = TRUE)
    p_val <- pchisq(G, df = 1, lower.tail = FALSE)
    
    data.frame(
      pair  = paste(p, collapse = " vs "),
      G     = round(G, 3),
      p_raw = round(p_val, 4)
    )
  })
  
  out       <- do.call(rbind, results)
  out$p_adj <- round(p.adjust(out$p_raw, method = "bonferroni"), 4)
  out
}

# Function to convert pairwise G-test results to CLD letters
gtest_to_cld <- function(pgtest_result) {
    pvals <- setNames(
    pgtest_result$p_adj,
    gsub(" vs ", "-", pgtest_result$pair))
  multcompView::multcompLetters(pvals)$Letters
}

# Format of p values in text, including the correct comparison operator and
# the italic "p" label. For values chained after a leading "*p*" (e.g.
# "*p* < 0.001, 0.011, 0.004, respectively"), set label = FALSE and
# symbol = FALSE so non-significant values print bare (no "="), while any
# value below 0.001 still automatically shows "< 0.001" regardless of
# position in the list.
format_pval <- function(p, label = TRUE, symbol = TRUE) {
  if (p < 0.001) {
    out <- "< 0.001"
  } else {
    val <- format(round(p, 3), nsmall = 3)
    out <- if (symbol) paste0("= ", val) else val
  }
  if (label) out <- paste0("*p* ", out)
  out
}

sig_label <- function(p) {
  case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ "ns"
  )
}

get_brms <- function(model, param, col = "Estimate") {
  tbl <- fixef(model) %>% as.data.frame() %>% rownames_to_column("parameter")
  round(tbl[[col]][tbl$parameter == param], 2)
}

format_brms <- function(model, param, post_prob = NULL) {
  b    <- get_brms(model, param)
  low  <- get_brms(model, param, "Q2.5")
  high <- get_brms(model, param, "Q97.5")
  base <- paste0("(\u03b2 = ", b, ", 95% CI: ", low, " to ", high)
  if (!is.null(post_prob)) {
    pp_label <- ifelse(round(post_prob, 2) >= 1, "> 0.99", as.character(round(post_prob, 2)))
    paste0(base, "; posterior probability = ", pp_label, ")")
  } else paste0(base, ")")
}

size_shape_map <- c(S = 15, M = 17, L = 18) 
sex_shape_map  <- c(F = 1, M = 16)

```


# Build DF's into long format
```{r}
#| label: build_dfs

# EXP 1 – Individual preference (long)
df_exp1_individual_long <- df_exp1_individual_raw %>%
  mutate(location_initial = initial, location_next_day = next_day) %>%
  pivot_longer(cols = matches("_(initial|next_day)$"),
    names_to = c(".value", "stage"),
    names_pattern = "^(.*)_(initial|next_day)$") %>%
  mutate(location_code = factor(toupper(location), levels = LOC_LEVELS),
    structure_type     = factor(structure, levels = STRUCTURE_LEVELS)) %>%
  select(tank, round, group_id, koura_id, size_class, sex, size_mm1, weight_g1, date, stage, t_leave_s, t_initial_s, location_code, structure_type, legs1, claws1, antenna1, size_mm2, weight_g2, legs2, claws2, antenna2,notes_exp, notes_koura)

write_out(df_exp1_individual_long, "df_exp1_individual_long.csv")

# EXP 2 – Group structure (long)
df_exp2_group_long <- df_exp2_group_raw %>%
  mutate(location_initial = initial, location_next_day = next_day) %>%
  pivot_longer(cols = c(location_initial, location_next_day,
      structure_initial, structure_next_day,legs_initial, claws_initial, antenna_initial,legs_next_day, claws_next_day, antenna_next_day), names_to = c(".value", "stage"), names_pattern = "^(.*)_(initial|next_day)$") %>%
  mutate(location_code  = factor(toupper(location), levels = LOC_LEVELS),
    structure_type = factor(structure, levels = STRUCTURE_LEVELS),
    size_mm_stage  = if_else(stage == "initial", size_mm1, size_mm2, NA_real_),
    weight_g_stage = if_else(stage == "initial", weight_g1, weight_g2, NA_real_)) %>%
  select(tank, round, group_id, koura_id, size_class, sex, date,stage, t_leave_s, t_initial_s, location_code, structure_type,legs, claws, antenna, size_mm_stage, weight_g_stage,size_mm1, weight_g1, legs1, claws1, antenna1,size_mm2, weight_g2, legs2, claws2, antenna2,notes_exp, notes_koura)

write_out(df_exp2_group_long, "df_exp2_group_long.csv")


# EXP 3 – Stone / wood (long)
df_exp3_woodstone_long <- df_exp3_woodstone_raw %>%
  mutate(location_initial = initial, location_next_day = next_day) %>%
  pivot_longer(cols = c(location_initial, location_next_day, structure_initial, structure_next_day),
    names_to = c(".value", "stage"),
    names_pattern = "^(.*)_(initial|next_day)$") %>%
  mutate(location_code = factor(toupper(location), levels = LOC_LEVELS3),
    structure_type = factor(structure, levels = STRUCTURE_LEVELS3)) %>%
  select(tank, round, group_id, koura_id, size_class, sex, date, stage, location_code, structure_type, size_mm1, weight_g1, legs1, claws1, antenna1, size_mm2, weight_g2, legs2, claws2, antenna2, notes_exp, notes_koura)

write_out(df_exp3_woodstone_long, "df_exp3_woodstone_long.csv")


# EXP 4 – Materials (long)
df_exp4_material_long <- df_exp4_material_raw %>%
  mutate(location_next_day = next_day) %>%
  pivot_longer(cols = c(location_next_day, structure_next_day),
    names_to = c(".value", "stage"),
    names_pattern = "^(.*)_(initial|next_day)$") %>%
  mutate(location_code = factor(toupper(location), levels = LOC_LEVELS),
    structure_type = factor(structure, levels = STRUCTURE_LEVELS4)) %>%
  select(tank, round, group_id, koura_id, size_class, sex, date, stage, location_code, structure_type,size_mm1, weight_g1, legs1, claws1, antenna1,size_mm2, weight_g2, legs2, claws2, antenna2,notes_exp, notes_koura)

write_out(df_exp4_material_long, "df_exp4_material_long.csv")
```

# Exp 1 Individual preference
## Exp 1 Statistic
```{r}
#| label: stats-exp1
#| tbl-cap: "Exp 1 statistical results"

# RQ 1: Is there a statistical difference between structure type in next_day?
# RQ 2: Is there a preference within each sex?
# RQ 3: Does structure type differ between sex?
# RQ 4: Did structure type use shift directionally between initial and next_day?

df1 <- df_exp1_individual_long %>%
  mutate(
    structure_type     = factor(structure_type),
    stage         = factor(stage, levels = c("initial", "next_day")),
    sex           = factor(sex),
    size_class    = factor(size_class),
    location_code = factor(location_code),
    tank          = factor(tank),
    round         = factor(round),
    koura_id      = factor(koura_id),
    size_mm1      = as.numeric(size_mm1),
    weight_g1     = as.numeric(weight_g1)) %>%
  filter(!is.na(structure_type))

size_stats1 <- df1 %>%
  distinct(koura_id, sex, size_mm1, weight_g1) %>%
  group_by(sex) %>%
  summarise(mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
    sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
    min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
    max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
    mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
    sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
    min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
    max_wt   = round(max(weight_g1,  na.rm = TRUE), 1),
    .groups  = "drop") %>%
  bind_rows(df1 %>%
      distinct(koura_id, size_mm1, weight_g1) %>%
      summarise(sex = "Total",
        mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
        sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
        min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
        max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
        mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
        sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
        min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
        max_wt   = round(max(weight_g1,  na.rm = TRUE), 1)))


df1_next_day <- df1 %>% filter(stage == "next_day" , structure_type != "TankWall") %>%
  mutate(structure_type = droplevels(structure_type))

pct_wood_overall <- round(mean(df1_next_day$structure_type == "WoodLog") * 100)
pct_wood_male    <- round(mean(df1_next_day$structure_type[df1_next_day$sex == "M"] == "WoodLog") * 100)
pct_wood_female  <- round(mean(df1_next_day$structure_type[df1_next_day$sex == "F"] == "WoodLog") * 100)
pct_walnext_day <- round(sum(df1$structure_type == "TankWall" & df1$stage == "next_day") / sum(df1$stage == "next_day") * 100)

# RQ1 — structure type preference at next_day
tab1_next_day <- table(df1_next_day$structure_type)
tab1_next_day_nz <- tab1_next_day[tab1_next_day > 0]

xmulti1 <- XNomial::xmulti(obs  = as.numeric(tab1_next_day_nz), expr = rep(sum(tab1_next_day_nz) / length(tab1_next_day_nz), length(tab1_next_day_nz)))
p_xmulti1 <- xmulti1$pLLR

# Post-hoc pairwise G-tests
pgtest1 <- pairwise_gtest(tab1_next_day_nz)
p_single_vs_flat1 <- pgtest1$p_adj[pgtest1$pair == "SingleStone vs FlatStone"]
p_single_vs_pile1 <- pgtest1$p_adj[pgtest1$pair == "SingleStone vs StonePile"]
p_single_vs_wood1 <- pgtest1$p_adj[pgtest1$pair == "SingleStone vs WoodLog"]
p_flat_vs_pile1   <- pgtest1$p_adj[pgtest1$pair == "FlatStone vs StonePile"]
p_flat_vs_wood1   <- pgtest1$p_adj[pgtest1$pair == "FlatStone vs WoodLog"]
p_pile_vs_wood1   <- pgtest1$p_adj[pgtest1$pair == "StonePile vs WoodLog"]

# Generate CLD letters for each panel
cld_combined1 <- data.frame(
  stage      = "next_day",
  panel      = "Combined",
  structure_type  = names(gtest_to_cld(pgtest1)),
  cld_letter = unname(gtest_to_cld(pgtest1)))

# RQ2 — Does sex affect structure preference?
# Model last fit: 2026-6-16
# Delete m_exp1_bayes.rds to refit
m_path1 <- file.path(der_data_dir, "m_exp1_bayes.rds")

if (!file.exists(m_path1)) {
  m_exp1_bayes <- brm(
    structure_type ~ sex + (1|round) + (1|tank),
    data    = df1_next_day,
    family  = categorical(),
    cores   = 4, iter = 4000, 
    open_progress = FALSE,
    control = list(adapt_delta = 0.99, max_treedepth = 12))
  saveRDS(m_exp1_bayes, m_path1)
} else {
  m_exp1_bayes <- readRDS(m_path1)
}

# Diagnostics
nuts_params(m_exp1_bayes) %>% 
  filter(Parameter == "divergent__") %>% 
  summarise(sum(Value))
summary(m_exp1_bayes)

# Posterior probability of sex effect
mean(as_draws_df(m_exp1_bayes)$b_muWoodLog_sexM > 0)


# RQ3 — Preference within each sex
structure_levels <- names(tab1_next_day_nz)

xmulti1_female <- XNomial::xmulti(
  obs  = as.numeric(table(df1_next_day$structure_type[df1_next_day$sex == "F"])),
  expr = rep(1/4, 4))
xmulti1_male <- XNomial::xmulti(
  obs  = as.numeric(table(df1_next_day$structure_type[df1_next_day$sex == "M"])),
  expr = rep(1/4, 4))

if (xmulti1_female$pLLR < 0.05) {
  pgtest1_female <- pairwise_gtest(table(df1_next_day$structure_type[df1_next_day$sex == "F"]))
  cld_female1    <- data.frame(stage = "next_day", panel = "Female",
    structure_type  = names(gtest_to_cld(pgtest1_female)),
    cld_letter = unname(gtest_to_cld(pgtest1_female)))} else {
  cld_female1 <- data.frame(stage = "next_day", panel = "Female",
    structure_type = structure_levels, cld_letter = "a")}

if (xmulti1_male$pLLR < 0.05) {
  pgtest1_male <- pairwise_gtest(table(df1_next_day$structure_type[df1_next_day$sex == "M"]))
  cld_male1    <- data.frame(stage = "next_day", panel = "Male",
    structure_type  = names(gtest_to_cld(pgtest1_male)),
    cld_letter = unname(gtest_to_cld(pgtest1_male)))} else {
  cld_male1 <- data.frame(stage = "next_day", panel = "Male",
    structure_type = structure_levels, cld_letter = "a")}

cld_alnext_day <- bind_rows(cld_combined1, cld_female1, cld_male1)

# RQ4 — Directional shift between initial and next_day
df1_wide     <- df1 %>%
  select(koura_id, stage, structure_type) %>%
  pivot_wider(names_from = stage, values_from = structure_type)

tab1_trans   <- table(initial = df1_wide$initial, next_day = df1_wide$next_day)
symtest1     <- coin::symmetry_test(tab1_trans, distribution = approximate(nresample = 9999))
p_symtest1   <- pvalue(symtest1)






```

## Exp 1 Figure
```{r}
#| label: fig-exp1
#| fig-width: 5
#| fig-height: 4
#| fig-cap: "Proportion of individual kōura (Experiment 1) found at each structure type the next day, shown separately for female (white bars) and male (black bars) kōura. Bars show observed proportions with sample sizes (n) shown on bars. Letters above bars indicate results of pairwise G-tests with Bonferroni correction within each sex group, structure types sharing the same letter within a panel do not differ significantly from each other, while structure types with different letters differ significantly in use. No significant difference in overall structure preference was detected between sexes (see results)."

exp1_structure_combined <- df_exp1_individual_long %>%
  filter(stage == "next_day") %>%
  count(stage, structure_type, name = "n") %>%
  group_by(stage) %>%
  mutate(N = sum(n), prop = n / N) %>%
  ungroup() %>%
  relevel_structure() %>%
  mutate(panel = "Combined")

exp1_structure_letters <- exp1_structure_combined %>%
  relevel_structure() %>%
  filter(stage == "next_day") %>%
  left_join(cld_alnext_day, by = c("stage", "panel", "structure_type"))

exp1_p_combined <- exp1_structure_combined %>%
  relevel_structure() %>%
  ggplot(aes(structure_type, prop)) +
  geom_col(color = "black") +
  geom_text(aes(label = n), vjust = 1.3, colour = "white", size = 3) +
  geom_text(data = exp1_structure_letters %>% filter(panel == "Combined"),
    aes(y = prop + 0.04, label = cld_letter), size = 3.5, fontface = "bold") +
  labs(x = "Structure type", y = "Proportion (next-day)")

exp1_structure_bysex <- df_exp1_individual_long %>%
  filter(stage == "next_day") %>%
  count(stage, structure_type, sex, name = "n") %>%
  group_by(stage, sex) %>%
  mutate(N = sum(n), prop = n / N) %>%
  ungroup() %>%
  relevel_structure() %>%
  mutate(panel = dplyr::recode(sex, F = "Female", M = "Male")) %>%
  select(-sex)

exp1_structure_letters_sex <- exp1_structure_bysex %>%
  left_join(cld_alnext_day, by = c("stage", "panel", "structure_type"))

exp1_p_sex <- exp1_structure_bysex %>%
  relevel_structure() %>%
  ggplot(aes(structure_type, prop, fill = panel)) +
  geom_col(color = "black", position = position_dodge(0.9)) +
  geom_text(aes(label = n), position = position_dodge(0.9), vjust = 1.3, colour = "grey50", size = 3) +
  geom_text(data = exp1_structure_letters_sex,  aes(y = prop + 0.04, label = cld_letter), position = position_dodge(0.9), size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c(Female = "white", Male = "black"), name = "Sex") +
  labs(x = "Structure type", y = "Proportion (next-day)") 

exp1_structure_p <- exp1_p_combined + exp1_p_sex 

ggsave(exp1_p_sex, file = file.path(out_dir, "fig-exp1.png"), width = 5, height = 4, dpi = 300)

exp1_p_sex
```


## Exp 1 II
```{r}
#| label: fig-exp1-tank-round
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "."

invisible(capture.output({
  # proportions by x var and facet var (per stage)
  mk_props <- function(data, xvar, facetvar) {
    data %>%
      dplyr::filter(stage %in% c("initial","next_day")) %>%
      dplyr::count(stage, {{ xvar }}, {{ facetvar }}, name = "n") %>%
      dplyr::group_by(stage, {{ facetvar }}) %>%
      dplyr::mutate(N = sum(n), prop = n / N) %>%
      dplyr::ungroup()
  }

  # bar plot; rows = facetvar, cols = stage; fixed y-scale across all plots
  mk_bar <- function(df, xvar, facetvar, title, structure = FALSE) {
    p <- ggplot(df, aes({{ xvar }}, prop)) +
      geom_col(color = "black") +
      geom_text(aes(label = n), vjust = 1.3, colour = "white") +
      facet_grid(rows = vars({{ facetvar }}), cols = vars(stage)) +
      labs(title = title) +
      scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05)))
    if (structure) {
      p <- p +
        scale_x_discrete(limits = STRUCTURE_LEVELS, drop = FALSE) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
    p
  }

  # data
  exp1_loc    <- mk_props(df_exp1_individual_long, location_code, tank)
  exp1_round  <- mk_props(df_exp1_individual_long, location_code, round)
  exp1_loc2   <- mk_props(df_exp1_individual_long, structure_type,     tank)
  exp1_round2 <- mk_props(df_exp1_individual_long, structure_type,     round)

  # plots
  p_exp1_loc    <- mk_bar(exp1_loc,   location_code, tank,  "Location by tank")
  p_exp1_round  <- mk_bar(exp1_round, location_code, round, "Location by round")
  p_exp1_loc2   <- mk_bar(exp1_loc2,  structure_type,     tank,  "Reeftype by tank",  structure = TRUE)
  p_exp1_round2 <- mk_bar(exp1_round2,structure_type,     round, "Reeftype by round", structure = TRUE)

  loc_round_plots_exp1 <<- p_exp1_loc + p_exp1_loc2 + p_exp1_round + p_exp1_round2 +
    patchwork::plot_layout(ncol = 4)
}))

ggsave(loc_round_plots_exp1, file = file.path(out_dir, "fig-exp1-tank-round.png"), width = 10, height = 7, dpi = 300)

loc_round_plots_exp1

```

## Exp 1 III
```{r}
#| label: fig-exp1-III
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "."

exp1_pairs <- df_exp1_individual_long %>%
  dplyr::filter(stage %in% c("initial","next_day")) %>%
  dplyr::select(koura_id, tank, round, size_class, sex, stage,location_code, structure_type, t_leave_s, t_initial_s) %>%
  dplyr::mutate(
    location_code = forcats::fct_drop(factor(location_code, levels = LOC_LEVELS)),
    structure_type     = forcats::fct_drop(factor(structure_type,     levels = STRUCTURE_LEVELS)),
    t_leave_s     = suppressWarnings(as.numeric(t_leave_s)),
    t_initial_s        = suppressWarnings(as.numeric(t_initial_s))) %>%
  tidyr::pivot_wider(names_from = stage,values_from = c(location_code, structure_type, t_leave_s, t_initial_s),names_sep = "_")

activity_df <- exp1_pairs %>%
  dplyr::transmute(koura_id, tank, round, size_class, sex, time_to_leave = t_leave_s_initial, time_at_initial = t_initial_s_initial) 

activity_long <- activity_df %>%
  mutate(
    time_to_leave = suppressWarnings(as.numeric(time_to_leave)),time_at_initial    = suppressWarnings(as.numeric(time_at_initial))) %>%
  pivot_longer(c(time_to_leave, time_at_initial), names_to = "metric", values_to = "time") %>%
  filter(!is.na(time), time >= 0)

lab_metric <- c(time_to_leave = "Time to leave start (t_leave)",
                time_at_initial    = "Time to reach initial hide (t_initial)")

act_exp1_p <- ggplot(activity_long, aes(time, col=sex)) +
  stat_ecdf(geom = "step") +
  facet_grid( ~ metric, labeller = labeller(metric = lab_metric)) +
  labs(x = "Time (s)", y = "Proportion completed by time")

act_exp1_p

```



# Exp 2 Group Dynamics
## Exp 2 Statistic
```{r}
#| label: stats-exp2
#| tbl-cap: "..."

# RQ 1: Does overall structure preference exist at group level? (Wilcoxon on group proportions)
# RQ 2: Does size class affect structure preference?            (brms categorical)
# RQ 3: Which structure is most preferred per size class?       (Posterior probabilities from brms)
# RQ 4: Did preference shift directionally initial→next_day?          (Stuart-Maxwell)

df2 <- df_exp2_group_long %>%
  mutate(
    structure_type  = factor(structure_type),
    stage      = factor(stage, levels = c("initial", "next_day")),
    sex        = factor(sex),
    size_class = factor(size_class, levels = c("S", "M", "L")),
    location_code = factor(location_code),
    tank       = factor(tank),
    round      = factor(round),
    group_id   = factor(group_id),
    koura_id   = factor(koura_id)) %>%
  filter(!is.na(structure_type))

size_stats2 <- df2 %>%
  distinct(koura_id, size_class, size_mm1, weight_g1) %>%
  group_by(size_class) %>%
  summarise(mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
    sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
    min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
    max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
    mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
    sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
    min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
    max_wt   = round(max(weight_g1,  na.rm = TRUE), 1),
    .groups  = "drop") %>%
  bind_rows(df2 %>%
      distinct(koura_id, size_mm1, weight_g1) %>%
      summarise(size_class = "Total",
        mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
        sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
        min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
        max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
        mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
        sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
        min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
        max_wt   = round(max(weight_g1,  na.rm = TRUE), 1)))

pct_wall2 <- round(sum(df2$structure_type == "TankWall" & df2$stage == "next_day") /
  sum(df2$stage == "next_day") * 100)


# RQ1 — Overall structure preference at group level
df2_grp <- df2 %>%
  filter(stage == "next_day") %>%
  count(group_id, round, structure_type) %>%
  complete(group_id, round, structure_type, fill = list(n = 0)) %>%
  group_by(group_id, round) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# Test whether proportion at each structure type differs from uniform (0.25)
wilcox_flat  <- wilcox.test(df2_grp$prop[df2_grp$structure_type == "FlatStone"],  mu = 0.25, exact = FALSE)
wilcox_pile  <- wilcox.test(df2_grp$prop[df2_grp$structure_type == "StonePile"],  mu = 0.25, exact = FALSE)
wilcox_wood  <- wilcox.test(df2_grp$prop[df2_grp$structure_type == "WoodLog"],    mu = 0.25, exact = FALSE)
wilcox_single <- wilcox.test(df2_grp$prop[df2_grp$structure_type == "SingleStone"], mu = 0.25, exact = FALSE)

p_wilcox2_adjusted <- p.adjust(
  c(p_wilcox_flat   = wilcox_flat$p.value,
    p_wilcox_single = wilcox_single$p.value,
    p_wilcox_pile   = wilcox_pile$p.value,
    p_wilcox_wood   = wilcox_wood$p.value),
  method = "bonferroni")
#p_wilcox2_adjusted

# Extract named values for inline use
p_wilcox2_flat   <- p_wilcox2_adjusted["p_wilcox_flat"]
p_wilcox2_single <- p_wilcox2_adjusted["p_wilcox_single"]
p_wilcox2_pile   <- p_wilcox2_adjusted["p_wilcox_pile"]
p_wilcox2_wood   <- p_wilcox2_adjusted["p_wilcox_wood"]

# Direct comparison FlatStone vs StonePile at group level
fs_props <- df2_grp %>% filter(structure_type == "FlatStone") %>% pull(prop)
sp_props <- df2_grp %>% filter(structure_type == "StonePile") %>% pull(prop)

wilcox_fs_vs_sp <- wilcox.test(fs_props, sp_props, paired = TRUE, exact = FALSE)
p_fs_vs_sp <- wilcox_fs_vs_sp$p.value
#p_fs_vs_sp


# RQ2 — Does size class affect structure preference?
# Model last fit: 2026-6-16
# Delete m2_size_bayes.rds to refit

# Path to saved model
m_path2 <- file.path(der_data_dir, "m_exp2_bayes.rds")

# Only fit if model file doesn't exist
if (!file.exists(m_path2)) {
  m_exp2_bayes <- brm(
    structure_type ~ size_class + (1 | group_id/koura_id) + (1 | round) + (1 | tank),
    data          = df2 %>% filter(stage == "next_day"),
    family        = categorical(),
    cores         = 4,
    iter          = 4000,
    open_progress = FALSE,
    control       = list(adapt_delta = 0.99, max_treedepth = 12))
  saveRDS(m_exp2_bayes, m_path2)
} else {
  m_exp2_bayes <- readRDS(m_path2)
}

# Diagnostics
nuts_params(m_exp2_bayes) %>% filter(Parameter == "divergent__") %>% summarise(sum(Value))
summary(m_exp2_bayes)

# Fixed effects summary
fixef(m_exp2_bayes)

# Posterior probability that medium preference for FlatStone is negative
p_post_flatM <- mean(as_draws_df(m_exp2_bayes)$b_muFlatStone_size_classM < 0)


# RQ3 — Which structure is most preferred per size class?
newdat <- data.frame(
  size_class = c("S", "M", "L"),
  group_id   = NA,
  koura_id   = NA,
  round      = NA)

pp <- posterior_epred(
  m_exp2_bayes,
  newdata          = newdat,
  re_formula       = NA,
  allow_new_levels = TRUE)

dimnames(pp)[[2]] <- c("S", "M", "L")
dimnames(pp)[[3]] <- head(STRUCTURE_LEVELS, dim(pp)[3])

pp_summary2 <- pp %>%
  as.data.frame.table(responseName = "prob") %>%
  rename(draw = Var1, size_class = Var2, structure_type = Var3) %>%
  group_by(size_class, structure_type) %>%
  summarise(
    mean_prob = mean(prob),
    lower     = quantile(prob, 0.025),
    upper     = quantile(prob, 0.975),
    .groups   = "drop")

# Posterior probability of being most preferred structure per size class
pct_top_S <- round(max(prop.table(table(apply(pp[,1,], 1, which.max)))) * 100)
pct_top_M <- round(max(prop.table(table(apply(pp[,2,], 1, which.max)))) * 100)
pct_top_L <- round(max(prop.table(table(apply(pp[,3,], 1, which.max)))) * 100)
pct_flat_L <- round(pp_summary2$mean_prob[pp_summary2$structure_type == "FlatStone" & pp_summary2$size_class == "L"] * 100)
pct_pile_L <- round(pp_summary2$mean_prob[pp_summary2$structure_type == "StonePile" & pp_summary2$size_class == "L"] * 100)


# RQ4 — Directional shift between initial and next_day
df2_wide <- df2 %>%
  filter(structure_type != "TankWall") %>%
  select(koura_id, group_id, round, stage, structure_type) %>%
  pivot_wider(names_from = stage, values_from = structure_type) %>%
  filter(!is.na(initial), !is.na(next_day)) %>%
  mutate(initial = factor(as.character(initial)), next_day = factor(as.character(next_day), levels = levels(initial)))

tab2_trans <- table(initial = df2_wide$initial, next_day = df2_wide$next_day)
smtest2    <- DescTools::StuartMaxwellTest(tab2_trans)
p_smtest2  <- smtest2$p.value

```

## Exp 2 Table
```{r}
#| label: tbl-exp2
#| tbl-cap: "Experiment 2 Bayesian model fixed effects"

fixef(m_exp2_bayes) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter") %>%
  rename(Mean = Estimate, SD = Est.Error,
         `2.5%` = Q2.5, `97.5%` = Q97.5) %>%
  knitr::kable(digits = 2)
```

## Exp 2 Figure
```{r}
#| label: fig-exp2
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Structure type preference of group kōura (Experiment 2). (A) Proportion of kōura found at each structure type at L1 across all groups and rounds. The dashed line indicates the null expectation of equal distribution (0.25). Significance symbols indicate deviation from null expectation (Wilcoxon signed-rank test, Bonferroni adjusted): ↑ used more than expected, ↓ used less than expected. Sample sizes shown in white. (B) Posterior predicted probability of structure type use by size class from Bayesian categorical mixed model, with 95 % credible intervals."

exp2_combined <- df_exp2_group_long %>%
  filter(stage == "next_day") %>%
  count(structure_type, name = "n") %>%
  mutate(N = sum(n), prop = n / N) %>%
  relevel_structure() %>%
  mutate(panel = "Combined")

# Wilcoxon significance
ann_wilcox2 <- data.frame(
  structure_type = c("SingleStone", "FlatStone", "StonePile", "WoodLog"),
  label     = c(sig_label(p_wilcox2_adjusted["p_wilcox_single"]),
                sig_label(p_wilcox2_adjusted["p_wilcox_flat"]),
                sig_label(p_wilcox2_adjusted["p_wilcox_pile"]),
                sig_label(p_wilcox2_adjusted["p_wilcox_wood"]))) %>%
  mutate(structure_type = factor(structure_type, levels = STRUCTURE_LEVELS)) %>%
  left_join(exp2_combined %>% select(structure_type, prop), by = "structure_type") %>%
  mutate(y = prop + 0.025, label = case_when(
           structure_type == "FlatStone" & label != "ns" ~ paste0(label, "↑"), structure_type %in% c("SingleStone", "WoodLog") & label != "ns" ~ paste0(label, "↓"), TRUE ~ label))

Exp2_combined_p <- exp2_combined %>%
  relevel_structure() %>%
  ggplot(aes(structure_type, prop)) +
  geom_col(color = "black", fill = "grey30") +
  geom_hline(yintercept = 0.25, linetype = "dashed", linewidth = 0.5) +
  geom_text(aes(label = n), vjust = 1.3, colour = "white") +
  geom_text(data = ann_wilcox2, aes(x = structure_type, y = y, label = label), size = 3.5, fontface = "bold") +
  #scale_y_continuous(limits = c(0, 0.7)) +
  labs(x = "Structure type", y = "Proportion (next-day)", tag = "A")

ce2_p <- pp_summary2 %>%
  filter(structure_type != "TankWall") %>%
  relevel_structure() %>%
  mutate(size_class = factor(size_class, levels = c("S", "M", "L"))) %>%
  ggplot(aes(structure_type, mean_prob, shape = size_class, group = size_class)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_shape_manual(values = size_shape_map, labels = c(S = "Small", M = "Medium", L = "Large")) +
  labs(x = "Structure type", y = "Predicted probability (95% CI)", shape = "Size class", tag = "B")


# Combine plots
Exp2_p <- Exp2_combined_p | ce2_p 

ggsave(Exp2_p, file = file.path(out_dir, "fig-exp2.png"), width = 8, height = 4, dpi = 300)

Exp2_p

```

## Exp 2 II
```{r}
#| label: fig-exp2-tank-round
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "."

invisible(capture.output({

  # proportions by x var and facet var (per stage)
  mk_props <- function(data, xvar, facetvar) {
    data %>%
      dplyr::filter(stage %in% c("initial","next_day")) %>%
      dplyr::count(stage, {{ xvar }}, {{ facetvar }}, name = "n") %>%
      dplyr::group_by(stage, {{ facetvar }}) %>%
      dplyr::mutate(N = sum(n), prop = n / N) %>%
      dplyr::ungroup()
  }

  # bar plot; uses facet_grid for rows=facetvar, cols=stage
  mk_bar <- function(df, xvar, facetvar, title, structure = FALSE) {
    p <- ggplot(df, aes({{ xvar }}, prop)) +
      geom_col(color = "black") +
      geom_text(aes(label = n), vjust = 1.3, colour = "white") +
      facet_grid(rows = vars({{ facetvar }}), cols = vars(stage)) +
      labs(title = title)+
      scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05)))
    if (structure) {
      p <- p +
        scale_x_discrete(limits = STRUCTURE_LEVELS, drop = FALSE) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
    p
  }

  # data
  exp2_loc    <- mk_props(df_exp2_group_long, location_code, tank)
  exp2_round  <- mk_props(df_exp2_group_long, location_code, round)
  exp2_loc2   <- mk_props(df_exp2_group_long, structure_type,     tank)
  exp2_round2 <- mk_props(df_exp2_group_long, structure_type,     round)

  # plots
  p_exp2_loc    <- mk_bar(exp2_loc,   location_code, tank,  "Location by tank")
  p_exp2_round  <- mk_bar(exp2_round, location_code, round, "Location by round")
  p_exp2_loc2   <- mk_bar(exp2_loc2,  structure_type,     tank,  "Reeftype by tank",  structure = TRUE)
  p_exp2_round2 <- mk_bar(exp2_round2,structure_type,     round, "Reeftype by round", structure = TRUE)

  loc_round_plots_exp2 <<- p_exp2_loc + p_exp2_loc2 + p_exp2_round + p_exp2_round2 +
    patchwork::plot_layout(ncol = 4)
}))

ggsave(loc_round_plots_exp2, file = file.path(out_dir, "fig-exp2-tank-round.png"), width = 10, height = 7, dpi = 300)

loc_round_plots_exp2

```

## Exp 2 III
```{r}
#| label: fig-exp2-III
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "."

exp2_structure_koura <- df_exp2_group_long %>%
  filter(stage == "next_day") %>%
  count(stage, structure_type, koura_id, name = "n") %>%
  group_by(stage, koura_id) %>%
  mutate(N = sum(n), prop = n / N) %>%
  ungroup() %>%
  relevel_structure()

p_exp2_structure_koura <- exp2_structure_koura %>%
  relevel_structure() %>%
  ggplot(aes(structure_type, prop)) +
  geom_col(color = "black") +
  facet_wrap(koura_id ~ stage, nrow = 6) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_exp2_structure_koura


exp2_use <- df_exp2_group_long %>% 
  filter(stage == "next_day")

counts <- exp2_use %>%
  filter(!is.na(structure_type), !is.na(size_class)) %>%
  mutate(
    tank = factor(tank),
    round = factor(round),
    stage = factor(stage),
    structure_type  = fct_relevel(structure_type,"SingleStone","FlatStone","StonePile","WoodLog","TankWall"),
    size_class = factor(size_class)) %>%
  count(stage, round, tank, structure_type, size_class, name = "number_koura") %>%
  complete(stage, round, tank, structure_type, size_class, fill = list(number_koura = 0))

p_stack <- ggplot(counts, aes(structure_type, number_koura, fill = size_class)) +
  geom_col() +
  facet_grid(round ~ tank+stage  , scales = "free_x")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_stack
```


# Exp 3 Wood vs stone
## Exp 3 Statistic
```{r}
#| label: stats-exp3
#| tbl-cap: "."

# RQ 1: Does overall structure preference exist at group level? (Wilcoxon on group proportions)
# RQ 2: Does size class affect structure preference?            (brms categorical)
# RQ 3: Which structure is most preferred per size class?       (Posterior probabilities from brms)
# RQ 4: Did preference shift directionally initial→next_day?          (Stuart-Maxwell)

df3 <- df_exp3_woodstone_long %>%
  mutate(
    structure_type      = factor(structure_type),
    stage          = factor(stage, levels = c("initial", "next_day")),
    sex            = factor(sex), # all male
    size_class     = factor(size_class, levels = c("S", "M", "L")),
    location_code  = factor(location_code),
    tank           = factor(tank),
    round          = factor(round),
    group_id       = factor(group_id),
    koura_id       = factor(koura_id)) %>%
  filter(!is.na(structure_type)) 

df3_next_day <- df3 %>% filter(stage == "next_day", !is.na(structure_type), structure_type != "NA")

pct_wall3 <- round(sum(df3$structure_type == "TankWall" & df3$stage == "next_day") / sum(df3$stage == "next_day") * 100)

# RQ1 — Overall structure preference at group level
df3_grp <- df3_next_day %>%
  count(group_id, round, structure_type) %>%
  filter(structure_type != "TankWall") %>%
  group_by(group_id, round) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

wilcox3_flat  <- wilcox.test(df3_grp$prop[df3_grp$structure_type == "FlatStone"], mu = 0.5, exact = FALSE)
wilcox3_split <- wilcox.test(df3_grp$prop[df3_grp$structure_type == "WoodSplit"], mu = 0.5, exact = FALSE)

p_wilcox3_adjusted <- p.adjust(c(p_wilcox3_flat  = wilcox3_flat$p.value,p_wilcox3_split = wilcox3_split$p.value), method = "bonferroni")

#p_wilcox3_adjusted

# Extract named values for inline use
p_wilcox3_flat  <- p_wilcox3_adjusted["p_wilcox3_flat"]
p_wilcox3_split <- p_wilcox3_adjusted["p_wilcox3_split"]


# RQ2 — Does size class affect structure preference?
# Model last fit: 2026-6-16
# Delete m_exp3_bayes.rds to refit

df3_next_day2 <- df3_next_day %>% 
  filter(structure_type != "TankWall") %>%
  mutate(structure_type = relevel(droplevels(factor(structure_type)), ref = "FlatStone"))

m_path3 <- file.path(der_data_dir, "m_exp3_bayes.rds")

if (!file.exists(m_path3)) {
  m_exp3_bayes <- brm(
    structure_type ~ size_class + (1 | group_id/koura_id) + (1 | round) + (1 | tank),
    data    = df3_next_day2,
    family  = bernoulli(),
    cores   = 4, iter = 4000, open_progress = FALSE,
    control = list(adapt_delta = 0.99, max_treedepth = 12))
  saveRDS(m_exp3_bayes, m_path3)
} else {
  m_exp3_bayes <- readRDS(m_path3)
}

# Diagnostics
nuts_params(m_exp3_bayes) %>% 
  filter(Parameter == "divergent__") %>% 
  summarise(sum(Value))
summary(m_exp3_bayes)
fixef(m_exp3_bayes)

# Posterior probability size class effects are negative (prefer FlatStone)
p_post3_M <- mean(as_draws_df(m_exp3_bayes)$b_size_classM < 0)
p_post3_L <- mean(as_draws_df(m_exp3_bayes)$b_size_classL < 0)

# RQ3 — Posterior predicted probability per size class
newdat3 <- data.frame(
  size_class = c("S", "M", "L"),
  group_id   = NA,
  koura_id   = NA,
  round      = NA,
  tank       = NA)

# pp3 is draws × 3: each value is P(WoodSplit) for that size class
pp3 <- posterior_epred(
  m_exp3_bayes,
  newdata          = newdat3,
  re_formula       = NA,
  allow_new_levels = TRUE)

colnames(pp3) <- c("S", "M", "L")

# Posterior probability FlatStone is preferred (P(WoodSplit) < 0.5)
pct_top3_S <- round(mean(pp3[, "S"] < 0.5) * 100)
pct_top3_M <- round(mean(pp3[, "M"] < 0.5) * 100)
pct_top3_L <- round(mean(pp3[, "L"] < 0.5) * 100)

# Mean predicted probability per structure per size class
pct_flat3_L  <- round((1 - mean(pp3[, "L"])) * 100)
pct_split3_L <- round(mean(pp3[, "L"]) * 100)

pp_summary3 <- data.frame(
  size_class     = rep(c("S", "M", "L"), each = 2),
  structure_type = rep(c("FlatStone", "WoodSplit"), 3),
  mean_prob      = c(1 - mean(pp3[,"S"]), mean(pp3[,"S"]),
                     1 - mean(pp3[,"M"]), mean(pp3[,"M"]),
                     1 - mean(pp3[,"L"]), mean(pp3[,"L"])),
  lower          = c(quantile(1 - pp3[,"S"], 0.025), quantile(pp3[,"S"], 0.025),
                     quantile(1 - pp3[,"M"], 0.025), quantile(pp3[,"M"], 0.025),
                     quantile(1 - pp3[,"L"], 0.025), quantile(pp3[,"L"], 0.025)),
  upper          = c(quantile(1 - pp3[,"S"], 0.975), quantile(pp3[,"S"], 0.975),
                     quantile(1 - pp3[,"M"], 0.975), quantile(pp3[,"M"], 0.975),
                     quantile(1 - pp3[,"L"], 0.975), quantile(pp3[,"L"], 0.975)))



# RQ4 — Directional shift initial → next_day
df3_wide <- df3 %>%
  filter(!is.na(structure_type), structure_type != "NA") %>%
  select(koura_id, group_id, round, stage, structure_type) %>%
  pivot_wider(names_from = stage, values_from = structure_type) %>%
  filter(!is.na(initial), !is.na(next_day)) %>%
  mutate( initial = factor(as.character(initial)), next_day = factor(as.character(next_day), levels = levels(initial)))

tab3_trans <- table(initial = df3_wide$initial, next_day = df3_wide$next_day)
smtest3    <- DescTools::StuartMaxwellTest(tab3_trans)
p_smtest3  <- smtest3$p.value

```

## Exp 3 Table
```{r}
#| label: tbl-exp3
#| tbl-cap: "Experiment 3 Bayesian model fixed effects"

fixef(m_exp3_bayes) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter") %>%
  rename(Mean = Estimate, SD = Est.Error,
         `2.5%` = Q2.5, `97.5%` = Q97.5) %>%
  knitr::kable(digits = 2)
```

## Exp 3 Figure
```{r}
#| label: fig-exp3
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Structure type preference in stone versus wood comparison (Experiment 3). (A) Proportion of kōura found at each structure type at next-day location across all groups and rounds. The dashed line indicates the null expectation of equal distribution (0.5). Significance symbols as in Figure 3. (B) Posterior predicted probability of structure type use by size class from Bayesian categorical mixed model, with 95 % credible intervals. ↑ used more than expected, ↓ used less than expected. TankWall excluded from model predictions."

exp3_combined <- df_exp3_woodstone_long %>%
  filter(stage == "next_day", !is.na(structure_type), structure_type != "NA") %>%
  count(structure_type, name = "n") %>%
  mutate(N = sum(n), prop = n / N) %>%
  relevel_structure3() %>%
  mutate(panel = "Combined")

ann_wilcox3 <- data.frame(
  structure_type = c("FlatStone", "WoodSplit"),
  label     = c(sig_label(p_wilcox3_adjusted["p_wilcox3_flat"]),
                sig_label(p_wilcox3_adjusted["p_wilcox3_split"]))) %>%
  mutate(structure_type = factor(structure_type, levels = STRUCTURE_LEVELS3)) %>%
  left_join(exp3_combined %>% select(structure_type, prop), by = "structure_type") %>%
  mutate(y = prop + 0.04, label = case_when( structure_type == "FlatStone" & label != "ns" ~ paste0(label, "↑"), structure_type %in% c("WoodSplit") & label != "ns" ~ paste0(label, "↓"), TRUE ~ label))

Exp3_combined_p <- exp3_combined %>%
  relevel_structure3() %>%
  ggplot(aes(structure_type, prop)) +
  geom_col(color = "black", fill = "grey30") +
  geom_hline(yintercept = .5, linetype = "dashed", linewidth = 0.5) +
  geom_text(aes(label = n), vjust = 1.3, colour = "white") +
  geom_text(data = ann_wilcox3, aes(x = structure_type, y = y, label = label), size = 3.5, fontface = "bold") +
  #scale_y_continuous(limits = c(0, 0.85)) +
  labs(x = "Structure type", y = "Proportion (next-day)", tag = "A")

ce3_p <- pp_summary3 %>%
  filter(structure_type != "NA", structure_type != "TankWall") %>%
  relevel_structure3() %>%
  mutate(size_class = factor(size_class, levels = c("S", "M", "L"))) %>%
  ggplot(aes(structure_type, mean_prob, shape = size_class, group = size_class)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_shape_manual(values = size_shape_map, labels = c(S = "Small", M = "Medium", L = "Large")) +
  labs(x = "Structure type", y = "Predicted probability (95% CI)", shape = "Size class", tag = "B")

Exp3_p <- Exp3_combined_p | ce3_p

ggsave(Exp3_p, file = file.path(out_dir, "fig-exp3.png"), width = 8, height = 4, dpi = 300)
Exp3_p
```

## Exp 3 II
```{r}
#| label: fig-exp3-tank-round
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "."

invisible(capture.output({
  # proportions by x var and facet var (per stage)
  mk_props <- function(data, xvar, facetvar) {
    data %>%
      dplyr::filter(stage %in% c("initial","next_day")) %>%
      dplyr::count(stage, {{ xvar }}, {{ facetvar }}, name = "n") %>%
      dplyr::group_by(stage, {{ facetvar }}) %>%
      dplyr::mutate(N = sum(n), prop = n / N) %>%
      dplyr::ungroup()
  }

# bar plot; uses facet_grid for rows=facetvar, cols=stage
  mk_bar <- function(df, xvar, facetvar, title, structure = FALSE) {
    p <- ggplot(df, aes({{ xvar }}, prop)) +
      geom_col(color = "black") +
      geom_text(aes(label = n), vjust = 1.3, colour = "white") +
      facet_grid(rows = vars({{ facetvar }}), cols = vars(stage)) +
      labs(title = title)+
      scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05)))
    if (structure) {
      p <- p +
        scale_x_discrete(limits = STRUCTURE_LEVELS3, drop = FALSE) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
    p
  }
  # data
  exp3_loc    <- mk_props(df_exp3_woodstone_long, location_code, tank)
  exp3_round  <- mk_props(df_exp3_woodstone_long, location_code, round)
  exp3_loc3   <- mk_props(df_exp3_woodstone_long, structure_type,     tank)
  exp3_round3 <- mk_props(df_exp3_woodstone_long, structure_type,     round)
  # plots
  p_exp3_loc    <- mk_bar(exp3_loc,   location_code, tank,  "Location by tank")
  p_exp3_round  <- mk_bar(exp3_round, location_code, round, "Location by round")
  p_exp3_loc3   <- mk_bar(exp3_loc3,  structure_type,     tank,  "Reeftype by tank",  structure = TRUE)
  p_exp3_round3 <- mk_bar(exp3_round3,structure_type,     round, "Reeftype by round", structure = TRUE)

loc_round_plots_exp3 <<- p_exp3_loc + p_exp3_loc3 + p_exp3_round + p_exp3_round3 +
    patchwork::plot_layout(ncol = 4)
}))

ggsave(loc_round_plots_exp3, file = file.path(out_dir, "fig-exp3-tank-round.png"), width = 10, height = 7, dpi = 300)

loc_round_plots_exp3
```


# Exp 4 Material comparison
## Exp 4 Statistic
```{r}
#| label: stats-exp4
#| tbl-cap: "."

# RQ 1: Does overall structure preference exist at group level? (Wilcoxon on group proportions)
# RQ 2: Does sex affect structure preference? (brms categorical)

df4 <- df_exp4_material_long %>%
  mutate(
    structure_type      = factor(structure_type),
    stage          = factor(stage, levels = c("next_day")),
    sex            = factor(sex),
    size_class     = factor(size_class, levels = c("S", "M", "L")),
    location_code  = factor(location_code),
    tank           = factor(tank),
    round          = factor(round),
    group_id       = factor(group_id),
    koura_id       = factor(koura_id)) %>%
  filter(!is.na(structure_type))

size_stats4 <- df4 %>%
  distinct(koura_id, sex, size_mm1, weight_g1) %>%
  group_by(sex) %>%
  summarise(mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
    sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
    min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
    max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
    mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
    sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
    min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
    max_wt   = round(max(weight_g1,  na.rm = TRUE), 1),
    .groups  = "drop") %>%
  bind_rows(df4 %>%
      distinct(koura_id, size_mm1, weight_g1) %>%
      summarise(sex = "Total",
        mean_ocl = round(mean(size_mm1,  na.rm = TRUE), 1),
        sd_ocl   = round(sd(size_mm1,    na.rm = TRUE), 1),
        min_ocl  = round(min(size_mm1,   na.rm = TRUE), 1),
        max_ocl  = round(max(size_mm1,   na.rm = TRUE), 1),
        mean_wt  = round(mean(weight_g1, na.rm = TRUE), 1),
        sd_wt    = round(sd(weight_g1,   na.rm = TRUE), 1),
        min_wt   = round(min(weight_g1,  na.rm = TRUE), 1),
        max_wt   = round(max(weight_g1,  na.rm = TRUE), 1)))

pct_wall4 <- round(sum(df4$structure_type == "TankWall" & df4$stage == "next_day") / sum(df4$stage == "next_day") * 100)

# RQ1 — Overall structure preference at group level
df4_grp <- df4 %>%
  count(group_id, round, structure_type) %>%
  filter(structure_type != "TankWall") %>%
  group_by(group_id, round) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# Test whether proportion at each structure type differs from uniform (.5)
wilcox_bart  <- wilcox.test(df4_grp$prop[df4_grp$structure_type == "ConcreteBrick"],  mu = .5, exact = FALSE)
wilcox_bwood  <- wilcox.test(df4_grp$prop[df4_grp$structure_type == "WoodBlock"],  mu = .5, exact = FALSE)


p_wilcox4_adjusted <- p.adjust(c(p_wilcox4_bart = wilcox_bart$p.value, p_wilcox4_bwood = wilcox_bwood$p.value), method = "bonferroni")

p_wilcox4_adjusted

# Extract named values for inline use
p_wilcox4_bart  <- p_wilcox4_adjusted["p_wilcox4_bart"]
p_wilcox4_bwood <- p_wilcox4_adjusted["p_wilcox4_bwood"]


# Direct comparison WoodBlock vs ConcreteBrick at group level
ba_props <- df4_grp %>% filter(structure_type == "ConcreteBrick") %>% pull(prop)
bw_props <- df4_grp %>% filter(structure_type == "WoodBlock") %>% pull(prop)


wilcox_bw_vs_ba <- wilcox.test(bw_props, ba_props, paired = TRUE, exact = FALSE)
p_bw_vs_ba <- wilcox_bw_vs_ba$p.value
p_bw_vs_ba

# RQ2 — Does sex affect structure preference?
# Model last fit: 2026-6-16
# Delete m_exp4_bayes.rds to refit
df4_model <- df4 %>%
  filter(stage == "next_day", structure_type != "TankWall") %>%
  mutate(structure_type = relevel(droplevels(factor(structure_type)), 
                                  ref = "ConcreteBrick"))

m_path4 <- file.path(der_data_dir, "m_exp4_bayes.rds")

if (!file.exists(m_path4)) {
  m_exp4_bayes <- brm(
    structure_type ~ sex + (1 | group_id/koura_id) + (1 | round) + (1 | tank),
    data    = df4_model,
    family  = bernoulli(),
    cores   = 4, iter = 4000, open_progress = FALSE,
    control = list(adapt_delta = 0.99, max_treedepth = 12))
  saveRDS(m_exp4_bayes, m_path4)
} else {
  m_exp4_bayes <- readRDS(m_path4)
}

# Diagnostics
nuts_params(m_exp4_bayes) %>% 
  filter(Parameter == "divergent__") %>% 
  summarise(sum(Value))
summary(m_exp4_bayes)

# Posterior probability sex effect (negative = prefer ConcreteBrick)
p_post_sexM <- mean(as_draws_df(m_exp4_bayes)$b_sexM < 0)

# Add tank to newdat4
newdat4 <- data.frame(sex= c("F", "M"), group_id = NA, koura_id = NA, round = NA,tank = NA)

# pp4 is now draws × 2 (F, M) — each value is P(WoodBlock)
pp4 <- posterior_epred(
  m_exp4_bayes,
  newdata          = newdat4,
  re_formula       = NA,
  allow_new_levels = TRUE)

colnames(pp4) <- c("F", "M")

pp_summary4 <- data.frame(
  sex            = rep(c("F", "M"), each = 2),
  structure_type = rep(c("ConcreteBrick", "WoodBlock"), 2),
  mean_prob      = c(1 - mean(pp4[,"F"]), mean(pp4[,"F"]),
                     1 - mean(pp4[,"M"]), mean(pp4[,"M"])),
  lower          = c(quantile(1 - pp4[,"F"], 0.025), 
                     quantile(pp4[,"F"], 0.025),
                     quantile(1 - pp4[,"M"], 0.025), 
                     quantile(pp4[,"M"], 0.025)),
  upper          = c(quantile(1 - pp4[,"F"], 0.975), 
                     quantile(pp4[,"F"], 0.975),
                     quantile(1 - pp4[,"M"], 0.975), 
                     quantile(pp4[,"M"], 0.975)))

pp_summary4

pct_bart4  <- round(mean(df4_grp$prop[df4_grp$structure_type == "ConcreteBrick"]) * 100)
pct_bwood4 <- round(mean(df4_grp$prop[df4_grp$structure_type == "WoodBlock"]) * 100)

```

## Exp 4 Table
```{r}
#| label: tbl-exp4
#| tbl-cap: "Experiment 4 Bayesian model fixed effects"

fixef(m_exp4_bayes) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter") %>%
  rename(Mean = Estimate, SD = Est.Error,
         `2.5%` = Q2.5, `97.5%` = Q97.5) %>%
  knitr::kable(digits = 2)
```

## Exp 4 Figure
```{r}
#| label: fig-exp4
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Material preference (Experiment 4). (A) Proportion of kōura found at each structure type at next-day location across all groups and rounds. The dashed line indicates the null expectation of equal distribution (0.5). (B) Posterior predicted probability of structure type use by sex from Bayesian categorical mixed model, with 95 % credible intervals. TankWall excluded from model predictions."

exp4_combined <- df_exp4_material_long %>%
  filter(stage == "next_day", !is.na(structure_type)) %>%
  count(structure_type, name = "n") %>%
  mutate(N = sum(n), prop = n / N) %>%
  relevel_structure4()

ann_wilcox4 <- data.frame(structure_type = c("ConcreteBrick", "WoodBlock"),label = c(sig_label(p_wilcox4_adjusted["p_wilcox4_bart"]), sig_label(p_wilcox4_adjusted["p_wilcox4_bwood"]))) %>%
  mutate(structure_type = factor(structure_type, levels = STRUCTURE_LEVELS4)) %>%
  left_join(exp4_combined %>% select(structure_type, prop), by = "structure_type") %>%
  mutate(y = prop + 0.04, label = case_when( structure_type %in% c("ConcreteBrick", "WoodBlock") & label != "ns" ~ paste0(label, "↑"), TRUE ~ label))

Exp4_combined_p <- exp4_combined %>%
  relevel_structure4() %>%
  ggplot(aes(structure_type, prop)) +
  geom_col(color = "black", fill = "grey30") +
  geom_hline(yintercept = .5, linetype = "dashed", linewidth = 0.5) +
  geom_text(aes(label = n), vjust = 1.3, colour = "white") +
  geom_text(data = ann_wilcox4, aes(x = structure_type, y = y, label = label), size = 3.5, fontface = "bold") +
  #scale_y_continuous(limits = c(0, 0.75)) +
  labs(x = "Structure type", y = "Proportion (next-day)", tag = "A")

ce4_p <- pp_summary4 %>%
  filter(structure_type != "TankWall") %>%
  relevel_structure4() %>%
  ggplot(aes(structure_type, mean_prob, shape = sex, group = sex)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_shape_manual(values = sex_shape_map, labels = c(F = "Female", M = "Male")) +
  labs(x = "Structure type", y = "Predicted probability (95% CI)", shape = "Sex", tag = "B")

Exp4_p <- Exp4_combined_p | ce4_p

ggsave(Exp4_p, file = file.path(out_dir, "fig-exp4.png"), width = 8, height = 4, dpi = 300)

Exp4_p
```

## Exp 4 II
```{r}
#| label: fig-exp4-tank-round
#| fig-width: 8
#| fig-height: 8
#| fig-cap: "."

df_exp4_material_long %>%
  group_by(group_id, size_class, sex, structure_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(group_id, size_class)

df_exp4_material_long %>%
  ggplot(aes(x = factor(group_id), fill = size_class)) +
  geom_bar(position = "fill") +  # use "stack" to show raw counts
  labs(x = "Group ID", y = "Proportion", fill = "Size class") +
  theme_bw()

df_exp4_material_long %>%
  ggplot(aes(x = factor(group_id), fill = interaction(sex, size_class))) +
  geom_bar(position = "fill") +
  labs(x = "Group ID", y = "Proportion", fill = "Sex × Size class") +
  theme_bw()

invisible(capture.output({
  # proportions by x var and facet var (per stage)
  mk_props <- function(data, xvar, facetvar) {
    data %>%
      dplyr::filter(stage %in% c("initial","next_day")) %>%
      dplyr::count(stage, {{ xvar }}, {{ facetvar }}, name = "n") %>%
      dplyr::group_by(stage, {{ facetvar }}) %>%
      dplyr::mutate(N = sum(n), prop = n / N) %>%
      dplyr::ungroup()
  }
  # bar plot; uses facet_grid for rows=facetvar, cols=stage
  mk_bar <- function(df, xvar, facetvar, title, structure = FALSE) {
    p <- ggplot(df, aes({{ xvar }}, prop)) +
      geom_col(color = "black") +
      geom_text(aes(label = n), vjust = 1.3, colour = "white") +
      facet_grid(rows = vars({{ facetvar }}), cols = vars(stage)) +
      labs(title = title)+
      scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05)))
    if (structure) {
      p <- p +
        scale_x_discrete(limits = STRUCTURE_LEVELS4, drop = FALSE) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
    p
  }

  # data
  exp4_loc    <- mk_props(df_exp4_material_long, location_code, tank)
  exp4_round  <- mk_props(df_exp4_material_long, location_code, round)
  exp4_loc4   <- mk_props(df_exp4_material_long, structure_type,     tank)
  exp4_round4 <- mk_props(df_exp4_material_long, structure_type,     round)

  # plots
  p_exp4_loc    <- mk_bar(exp4_loc,   location_code, tank,  "Location by tank")
  p_exp4_round  <- mk_bar(exp4_round, location_code, round, "Location by round")
  p_exp4_loc4   <- mk_bar(exp4_loc4,  structure_type,     tank,  "Reeftype by tank",  structure = TRUE)
  p_exp4_round4 <- mk_bar(exp4_round4,structure_type,     round, "Reeftype by round", structure = TRUE)

  loc_round_plots_exp4 <<- p_exp4_loc + p_exp4_loc4 + p_exp4_round + p_exp4_round4 +
    patchwork::plot_layout(ncol = 4)
}))

ggsave(loc_round_plots_exp4, file = file.path(out_dir, "fig-exp4-tank-round.png"), width = 10, height = 7, dpi = 300)

loc_round_plots_exp4

```

# Combine figure for presentation
```{r}
#| label: fig-all-presentation
#| fig-width: 7
#| fig-height: 7
#| fig-cap: "."

p1 <- exp1_p_combined + labs(x = NULL, y = "Proportion", title = "Exp 1: Single kōura")
p2 <- Exp2_combined_p + labs(x = NULL, y = NULL, title = "Exp 2: Group kōura", tag = NULL)
p3 <- Exp3_combined_p + labs(x = NULL, y = "Proportion", title = "Exp 3: Wood vs Stone", tag = NULL)
p4 <- Exp4_combined_p + labs(x = NULL, y = NULL, title = "Exp 4: Bricks", tag = NULL)

fig_all_presentation <- p1 + p2 + p3 + p4 + plot_layout(nrow = 2)

ggsave(fig_all_presentation, file = file.path(out_dir, "fig-all-presentation.png"), width = 6.5, height = 6.5, dpi = 300)
fig_all_presentation

```

```{r}
#| label: fig-ce-all-presentation
#| fig-width: 7.5
#| fig-height: 7
#| fig-cap: "."


sex1_p <- exp1_p_sex + labs(x = NULL, y = "Proportion (next_day)", title = "Exp 1: Single kōura by sex") + theme(legend.position = "none")
ce2_pp <- ce2_p + labs(x = NULL, y = "Predicted probability (95% CI)", title = "Exp 2: Group kōura by size", tag = NULL)
ce3_pp <- ce3_p + labs(x = NULL, y = "Predicted probability (95% CI)", title = "Exp 3: Wood vs Stone by size", tag = NULL) + theme(legend.position = "none")
ce4_pp <- ce4_p + labs(x = NULL, y = "Predicted probability (95% CI)", title = "Exp 4: Bricks by sex", tag = NULL)

ce_all <- sex1_p + ce2_pp + ce3_pp + ce4_pp + plot_layout(nrow = 2)

ggsave(ce_all, file = file.path(out_dir, "fig-ce-all-presentation.png"), width = 7.5, height = 7, dpi = 300)

ce_all
```

# Waterquality
```{r}
#| label: fig-wq
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "."

wq_clean <- df_wq %>%
  mutate(value = suppressWarnings(as.numeric(value)),tank  = factor(tank),param_label = ifelse(is.na(unit) | unit %in% c("", "NA"), parameter_abb, paste(parameter_abb, unit))) %>%
  filter(!is.na(tank_use), !is.na(value),data_time >= as.POSIXct("2025-09-01"),tank_use == "Experiment")

wq_summary <- wq_clean %>%
  filter(param_label %in% c("Temp (°C)", "DO (%)")) %>%
  group_by(param_label) %>%
  summarise(
    mean_val = round(mean(value, na.rm = TRUE), 1),
    sd_val   = round(sd(value, na.rm = TRUE), 1),
    min_val  = round(min(value, na.rm = TRUE), 1),
    max_val  = round(max(value, na.rm = TRUE), 1),
    .groups  = "drop")

mean_temp <- wq_summary$mean_val[wq_summary$param_label == "Temp (°C)"]
sd_temp   <- wq_summary$sd_val[wq_summary$param_label == "Temp (°C)"]
mean_do   <- wq_summary$mean_val[wq_summary$param_label == "DO (%)"]
sd_do     <- wq_summary$sd_val[wq_summary$param_label == "DO (%)"]




wq_probe <- wq_clean %>%
  transmute(datetime = data_time,tank,param_label,value)

# --- Logger data (df_loggers) -> give unique param_label names ---
wq_logger <- df_loggers %>%
  rename(datetime = data_time) %>%
  mutate(tank     = factor(tank),
    datetime = lubridate::ymd_hms(datetime, tz = "Pacific/Auckland")) %>%
  pivot_longer(c(temperature, light), names_to = "param", values_to = "value") %>%
  mutate(
    param_label = case_when(
      param == "temperature" ~ "Temp (°C) - Logger",
      param == "light"       ~ "Light (lux) - Logger",
      TRUE                   ~ param),
    date  = as.Date(datetime),
    time  = format(datetime, "%H:%M:%S"),
    year  = year(datetime),
    month = month(datetime, label = TRUE),
    day   = day(datetime)) %>%
  filter(
    datetime >= as.POSIXct("2025-09-01", tz = "Pacific/Auckland"),
    (param_label != "Light (lux) - Logger" | value <= 1000),
    (param_label != "Temp (°C) - Logger"   | value <= 20) ) %>%
  transmute(datetime, date, time, year, month, day, tank, param_label, value)

# --- Combine into ONE df, keep labels distinct ---
PARAM_LEVELS <- c("Temp (°C)", "Temp (°C) - Logger","Light (lux) - Logger",
                  "Press (mbars)","DO (%)","DO (mg/L)","SPC (µS/cm)","Cond (µS/cm)","SAL (ppt)","pH")

wq_all <- bind_rows(wq_probe, wq_logger) %>%
  mutate(param_label = factor(param_label, levels = PARAM_LEVELS),tank = factor(tank)) %>%
  tidyr::drop_na(value)

# --- Plot from the single df ---
wq_all_plot <- ggplot(wq_all, aes(datetime, value, color = tank, group = tank)) +
  geom_line() +
  facet_wrap(~ param_label, scales = "free_y", ncol = 2)

wq_all_plot
ggsave(wq_all_plot, file = file.path(out_dir, "fig-wq.png"), width = 5, height = 5, dpi = 300)

```


# Length-Weight of koura
```{r}
#| label: fig-length-weight
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "."

df_size_weight_all <- dplyr::bind_rows(
  df_exp1_individual_raw %>%
    dplyr::transmute(experiment = "Exp 1: Single kōura",
      size_mm1   = as.numeric(size_mm1),
      weight_g1  = as.numeric(weight_g1),
      sex= as.factor(sex)), df_exp2_group_raw %>%
    dplyr::transmute(
      experiment = "Exp 2: Group kōura",
      size_mm1   = as.numeric(size_mm1),
      weight_g1  = as.numeric(weight_g1),
      sex= as.factor(sex)),df_exp3_woodstone_raw %>%
    dplyr::transmute(
      experiment = "Exp 3: Wood vs Stone",
      size_mm1   = as.numeric(size_mm1),
      weight_g1  = as.numeric(weight_g1),
      sex= as.factor(sex)),df_exp4_material_raw %>%
    dplyr::transmute(
      experiment = "Exp 4: Bricks",
      size_mm1   = as.numeric(size_mm1),
      weight_g1  = as.numeric(weight_g1),
      sex= as.factor(sex))) %>%
  dplyr::filter(!is.na(size_mm1), !is.na(weight_g1)) %>%
  dplyr::mutate(experiment = factor(experiment))

p_size_weight <- ggplot2::ggplot(
  df_size_weight_all,
  ggplot2::aes(x = size_mm1, y = weight_g1, colour = sex)) +
  ggplot2::geom_vline(xintercept = 22, linetype = "dotted") +
  ggplot2::geom_vline(xintercept = 30, linetype = "dotted") +
  ggplot2::geom_point()+
  ggplot2::facet_grid(~ experiment)+
  labs(x = "OCL Length mm", y = "Weight g")

p_size_weight

```


# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```